---
title: "Midterm Review"
author: "CSCI 36: Foundations of Data Science"
date: "2/22/2021"
output:
  html_document:
    css: midterm-review-style.css
  pdf_document: default
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, results='hide'}
library("tidyverse")
library("nycflights13")
```

**Knit me to html for best results!!!**

### Useful Resources (Click Me!!!)

##### Textbook
- [Foundations of Data Science](https://piazza.com/class_profile/get_resource/kk4x4g7te674nc/kk8xkp4od685jm)

##### Homework Solutions
- [Homework 1 Solutions](https://piazza.com/class_profile/get_resource/kk4x4g7te674nc/klabgv35he945y)
- [Homework 2 Solutions](https://piazza.com/class_profile/get_resource/kk4x4g7te674nc/klabh4im9av4h8)
- [Homework 3 Solutions](https://piazza.com/class_profile/get_resource/kk4x4g7te674nc/klabhdck3ze4sf)
- [Homework 4 Solutions](https://piazza.com/class_profile/get_resource/kk4x4g7te674nc/kldwfuzwn9k93)

##### Lectures
- [Lecture 1: Introduction](https://www.youtube.com/watch?v=bj3jZHc7aO4&feature=youtu.be)
- [Lecture 2: Tidyverse and Select](https://www.youtube.com/watch?v=dYtSLu40D9U&feature=youtu.be)
- [Lecture 3: Filter, Arrange, and Mutate](https://www.youtube.com/watch?v=RwlZUsPZVT4&feature=youtu.be)
- [Lecture 4: Group-by and Aggregation Functions](https://www.youtube.com/watch?v=SIXv30cobHU&feature=youtu.be)
- [Lecture 5: Group-by and Summarise](https://www.youtube.com/watch?v=DLpWHBjiEJw&feature=youtu.be)
- [Lecture 6: Tidy Data](https://www.youtube.com/watch?v=qrY-7rvwAuY&feature=youtu.be)
- [Lecture 7: Data Types](https://www.youtube.com/watch?v=WVAWzKLRMkc&feature=youtu.be)
- [Lecture 8: Midterm Review](https://www.youtube.com/watch?v=i8DrZQn0hJo&feature=youtu.be)

### **Aggregation Functions**
[For more information click here!](https://www.guru99.com/r-aggregate-function.html#3)
```{r, results='hide'}
x <- c(1,2,3)
mean(x)        # Average of vector
median(x)      # Median of vector
sum(x)         # Sum of vector
sd(x)          # Standard deviation of vector
IQR(x)         # Interquartile range of vector
min(x)         # Minimum of vector
max(x)         # Maximum of vector
quantile(x)    # Quantile of vector
first(x)       # First observation of the group
last(x)        # Last observation of the group
nth(x,1)       # nth observation of the group
# n()          # Count the number of rows
n_distinct(x)  # Count the number of distinct observations
```

### **Select**
Select (and optionally rename) variables in a data frame by referring to them by name.

##### **Constructor**
- `select(.data, ...)`

##### **Arguments**
- `.data`: A data frame.
- `...`: Specification of the desired column(s).  One or more unquoted expressions separated by commas.  Variable names can be used as if they were positions in the data frame, so expressions like x:y canbe used to select a range of variables.

##### **Helpers**
- `:` for selecting a range of consecutive variables
- `!/-` for taking the complement of a set of variables
- `&` for selecting the intersection of two sets of variables
- `|` for selecting the union of two sets of variables
- `c()` for combining selections
- `everything()` Matches all variables
- `last_col()` Select last variable, possibly with an offset
- `starts_with()` Starts with a prefix
- `ends_with()` Ends with a suffix
- `contains()` Contains a literal string
- `matches()` Matches a regular expression
- `num_range()` Matches a numerical range
- `all_of()` Matches variable names in a character vector.  All names must be present, otherwise an out-of-bounds error is thrown
- `any_of()` Same as all_of() except that no error is thrown for anmes that don't exist
- `where()` Applies a function to all variables and selects those for which the function returns True

##### **Examples**
1. Select the carrier and tail number from the flights datas et
```{r, results='hide'}
planes <- select(flights, carrier, tailnum)
planes
```

2. Select all variables except the year from the flights data set
```{r, results='hide'}
not_year1 <- select(flights, -year)
not_year2 <- select(flights, !year)
not_year1
```

3. Select all variable ending with the word "time" from the flights data set
```{r, results='hide'}
time_variables <- select(flights, ends_with("time"))
time_variables
```

4. Select the first 9 variables from the flights data set
```{r, results='hide'}
first_nine <- select(flights, 1:9)
first_nine
```


### **Filter**
Used to subset a data frame, retaining rows that satisfied all conditions (i.e. evaluating to True). Can be applied to both grouped and un-grouped data.

##### **Constructor**
- `filter(.data, ..., .preserve = T/F)`

##### **Arguments**
- `.data`: A data frame.
- `...`: Expressions that return a logical value (i.e. True or False) and are defined in terms of variables in `.data`.  If multiple expressions are included, they are combined with the `&` operator.
- `.preserve`: Relevant when the `.data` input is grouped.  If `.preserve = False` (the default), the grouping structure is recalculated based on the resulting data, otherwise the grouping is kept as is. 

##### **Helpers**
- `==`, `!=`, `>`, `<`, `>=`, `<=`
- `&`, `|`, `xor()`, `!`
- `is.na()`
- `between()`, `near()`
- `abs()`

##### **Examples**
1. Filter for all flights with arrival delay of two or more hours from the flights data set
```{r, results='hide'}
arrival_delay <- filter(flights, arr_delay >= 120)
arrival_delay
```

2. Filter for all flights that flew to Houston (airport codes "IAH" or "HOU") from the flights data set
```{r results='hide'}
to_houston <- filter(flights, dest == "IAH" | dest == "HOU")
to_houston
```

3. Filter for all flights that departed from JFK in July from the flights data set
```{r, results='hide'}
JFK_July1 <- filter(flights, origin == "JFK" & month == 7)
JFK_July2 <- filter(flights, origin == "JFK", month == 7)
JFK_July1
```

4. Filter for all flights that left between 0 and 60 minutes late from the flights data set
```{r results='hide'}
late <- filter(flights, between(dep_delay, 0, 60))
late
```

5. Filter for all flights that do not have missing departure times from the flights data set
```{r results='hide'}
flights_with_dep_time <- filter(flights, !is.na(dep_time))
flights_with_dep_time
```

### **Arrange**
Orders the rows of a data frame by the values of selected columns. NA values are always put at the end.

##### **Constructor**
- `arrange(.data, ..., .by_group = T/F)`

##### **Arguments**
- `.data`: A data frame.
- `...`: Variables or functions (Note: use `desc()` to sort a variable in descending order).
- `.by_group`: If True, will sort by grouping variable.

##### **Helpers**
- `desc()` arrange in ascending order
- `abs()`

##### **Examples**
1. Arrange the flights data set so that the flights that departed closest to their scheduled departure time are first
```{r, results='hide'}
most_on_time <- arrange(flights, abs(dep_delay))
most_on_time
```

2. Arrange the flights data set from longest to shortest arrival delay
```{r, results='hide'}
latest_arrival <- arrange(flights, desc(arr_delay))
latest_arrival
```

### **Mutate**
Adds new variables and preserves existing one. New variables overwrite existing variables of the same name.

##### **Constructor**
- `mutate(.data, ...)`

##### **Arguments**
- `.data`: A data frame.
- `...`: Name-value pairs.  The name gives the name of the column in the output.  The value can be: a vector of length 1, a vector the same length as the current group, null, or a data frame or tibble.

##### **Helpers**
- `+`, `-`, `*`. `/`
- `log()`
- `==`, `!=`, `>`, `<`, `>=`, `<=`    For logical results
- `&`, `|`, `xor()`, `!`              For logical results

##### **Examples**
1. Use mutate to convert the distance variable, in the flights data set, from miles to feet 
```{r results='hide'}
flights_with_feet <- mutate(flights, distance = distance * 5280)
flights_with_feet
```

2. Use mutate to add a variable "speed" to the "flights" table that gives the average flight speed, in mph
```{r results='hide'}
flights_with_speed <- mutate(flights, speed = distance / (air_time/60))
flights_with_speed
```

3. Use mutate to add a variable "early" which is True if the flight arrived early and False if it arrived on time or late
```{r results='hide'}
early_flights <- mutate(flights, early = arr_delay < 0)
early_flights
```

### **Group By**
Takes an existing table and converts it into a grouped table where operations are performed "by group"

##### **Constructor**
- `group_by(.data, ..., .add)`

##### **Arguments**
- `.data`: A data frame.
- `...`: Variables or computations to group by
- `.add`: When False, the default, group_by() will override existing groups. To add to existing groups, use .add = True

##### **Examples**
1. Use group_by to determine if all characters from a given homeworld all have brown eyes from the starwars data set.
```{r, results='hide'}
brown <- starwars %>%
  group_by(homeworld) %>%
  summarise(all_brown = all(eye_color == "brown"), number = sum(eye_color == "brown"))
brown
```

2. Use group-by to determine the most common eye colors among star wars characters from the starwars data set.
```{r, results='hide'}
eyes <- starwars %>%
  group_by(eye_color) %>%
  summarise(number = n())
eyes
```

### **Ungroup**
Takes an existing table with groups and removes grouping

##### **Constructor** 
- `ungroup(x, ...)`

##### **Arguments**
- `x`: A tibble
- `...`: Variables to remove from grouping

##### **Examples**
1. Use group_by to group the data by eye color and then use un-group to get back the original data from the starwars data set.
```{r, results='hide'}
ungroup_eyes <- starwars %>%
  group_by(eye_color) %>%
  ungroup(eye_color)
ungroup_eyes
```

### **Summarise**
Creates a new data frame.  It will have one (or more) rows for each combination of grouping variables; if there are no grouping variables, the output will have a single row summarizing all observations in the input.  It will contain one column for each grouping variable and one column for each of the summary statistics that you have specified.

##### **Constructor**
- `summarise(.data, ...)`

##### **Arguments**
- `.data`: A data frame.
- `...`: Name-value pairs of summary functions.  The name will be the name of the variable in the result.  The value can be a vector of length 1, a vector of length n, or a data frame

##### **Helpers**
- `mean()`, `median()`            Center
- `sd()`, `IQR()`, `mad()`        Spread
- `min()`, `max()`, `quartile()`  Range
- `first()`, `last()`, `nth()`    Position
- `n()`, `n_distinct()`           Count
- `any()`, `all()`                Logical

##### **Examples**
1. Use the summarize function to calculate the minimum and maximum mass among all characters of the same sex from the starwars data set.
```{r, results='hide'}
sex_mass <- starwars %>%
  group_by(sex) %>%
  summarise(min_mass = min(mass, na.rm = TRUE), max_mass = max(mass, na.rm = TRUE))
sex_mass
```

2. Use the summarize function to calculate the average heights of the three shortest character form each homeworld in the starwars data set.
```{r, results='hide'}
shortest_3_avg <- starwars %>%
  group_by(homeworld) %>%
  filter(n() >= 3, rank(height) <= 3) %>%
  summarise(avg = sum(height) / 3)
shortest_3_avg
```

### **Pivot Wider**
"Widens" the data, increasing the number of columns and decreasing the number of rows

##### **Constructor**
- `pivot_wider(
  data,
  id_cols = NULL,
  names_from = name,
  names_prefix = "",
  names_sep = "_",
  names_glue = NULL,
  names_sort = FALSE,
  names_repair = "check_unique",
  values_from = value,
  values_fill = NULL,
  values_fn = NULL,
  ...
)`

##### **Arguments**
- `data`: A data frame to pivot.
- `names_from`, `values_from`: A pair of arguments describing which column (or columns) to get the name of the output column (`names_from`), and which column (or columns) to get the cell values from (`values_from`)

##### **Examples**
1. Use pivot wider to 
```{r, results='hide'}
us_rent_income
tidy_rent <- us_rent_income %>%
  select(-GEOID, -moe) %>%
  pivot_wider(names_from = variable, values_from = estimate)
tidy_rent
```


### **Pivot Longer**
"Lengthens" the data, increasing the number of rows and decreasing the number of columns.

##### **Constructor**
- `pivot_longer(
  data,
  cols,
  names_to = "name",
  names_prefix = NULL,
  names_sep = NULL,
  names_pattern = NULL,
  names_ptypes = list(),
  names_transform = list(),
  names_repair = "check_unique",
  values_to = "value",
  values_drop_na = FALSE,
  values_ptypes = list(),
  values_transform = list(),
  ...
)`

##### **Arguments**
- `data`: A data frame to pivot
- `cols`: Columns to pivot into longer format
- `names_to`: A string specifying the name of the column to create from the data stored in the column names of `data`.
- `values_to`: A string specifying the name of the column to create from the data stored in cell values.

##### **Examples**
1. Use pivot longer to consolidate the multiple month columns into a single month column.
```{r, results='hide'}
rainfall <- read_csv2('City;January;February;March
  Atlanta, Georgia;11;10;10
  Austin, Texas;7;7;9
  Baltimore, Maryland;10;9;10
  Birmingham, Alabama;11;10;10
  Boston, Massachusetts;11;10;12')
rainfall

month_column <- rainfall %>%
  pivot_longer(cols = c("January", "February", "March"), names_to = "Month", values_to = "Rainfall")
month_column
```

### **Separate**
Given either a regular expression or a vector of character positions, `separate()` turns a single character column into multiple columns.

##### **Constructor**
- `separate(
  data,
  col,
  into,
  sep = "[^[:alnum:]]+",
  remove = TRUE,
  convert = FALSE,
  extra = "warn",
  fill = "warn",
  ...
)`

##### **Arguments**
- `data`: A data frame.
- `col`: Column name or position.
- `into`: Names of new variables to create as character vector.
- `sep`: Separator between columns. If character, `sep` is interpreted as a regular expression.  The default values is a regular expression that matches any sequence of non-alphanumeric values.  If numeric, `sep` is interpreted as character positions to split at. Positive values start at 1 at the far-left of the string; negative values start at -1 at the far-right of the string

##### **Examples**
1. Use separate to split up the City column into city and state columns.
```{r, results='hide', warning=FALSE, message=FALSE}
rainfall <- read_csv2('City;January;February;March
  Atlanta, Georgia;11;10;10
  Austin, Texas;7;7;9
  Baltimore, Maryland;10;9;10
  Birmingham, Alabama;11;10;10
  Boston, Massachusetts;11;10;12')
rainfall

separate_rainfall <- rainfall %>%
  separate(col = "City", into = c("City", "State"), sep = ",")
separate_rainfall
```

### **Unite**
Convenience function to paste together multiple columns into one

##### **Constructor** 
- `unite(data, col, ..., sep = "_", remove = TRUE, na.rm = FALSE)`

##### **Arguments**
- `data`: A data frame.
- `col`: The name of the new column, as a string or symbol.
- `...`: Columns to unite.
- `sep`: Separator to use between values.
- `remove`: If True, remove input columns form output data frame.
- `na.rm`: If True, missing values will be removed prior to uniting each value

##### **Examples**
1. Use unite to combine the month, day, and year columns into a date column.
```{r, results='hide', warning=FALSE, message=FALSE}
dates <- read_csv2('Month;Day;Year
  3;2;2001
  4;7;2003
  6;15;2004
  9;30;2007
  8;1;2015'
)
dates

unite_date <- dates %>%
  unite(Date, Month, Day, Year, sep = "-")
unite_date
```